---
title: 'Course Project: Practical ML'
author: "Vadim Sovetkin"
date: "2026-04-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.path = "figs/"
)
```

```{r libraries, echo=FALSE}
library(tidyverse)
library(caret)
library(doParallel)
```

## 1. Load Data

The training data were pre-downloaded from here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv: 

- The number of features in the original training data set is 159, plus the target variable. 

The testing data were pre-downloaded from here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv: 

- The number of features in the original testing data set is 159, plus the problem ID variable. 

```{r read_data, echo=TRUE}

# Read data from CSV file 
training <- read.csv("pml-training.csv", stringsAsFactors = FALSE)
testing <- read.csv("pml-testing.csv", stringsAsFactors = FALSE)

# Dimensions 
cat("Training set dimensions:", dim(training))
cat("\nTesting set dimensions: ", dim(testing))

```
The goal of the project is to predict the **classe** variable in the target and use any of the other variables as predictors: 

- The variable **class** was converted from character type to a factor with 5 levels. 

The distribution of the **classe** target variable was analysed

- The frequency for for class A at around 28% is higher than for other classes; 
- The frequency the classes B-E is fairly uniform in the range 16-19%. 

```{r examine_target, echo=TRUE}

# Convert char to factor 
training$classe <- factor(training$classe)

# Show a frequency table 
cat("Proportions of the target classes:\n")
round( prop.table(table(training$classe, useNA="ifany")), 3)

```

## 2. Examine Data Features 

Data exploration and pre-processing was focused on: 

- Identifying and removing empty or uninformative predictors; 
- Assessing co-linearity (correlation) between predictors; 
- Decorrelating and reducing the number of features using PCA. 


### 2.1. Remove empty features 

A total of 100 features were found to have less that 10% non-empty values: 

- These mostly-empty features were removed from both training and test data sets. 

```{r remove_NA, echo=TRUE}

# Replace #DIV/0! with NA
training[ training == "#DIV/0!" ] <- NA
training[ training == "" ] <- NA

# Identify mostly empty columns 
min_count <- dim(training)[2] * 0.90
counts_NA <- training %>% summarise( across(everything(), ~ sum(is.na(.))) )
cat("Number of mostly empty features (NAs > 90%): ", sum(counts_NA > min_count))

# Remove mostly empty columns 
features_select <- names( counts_NA[ which(counts_NA < min_count) ] )
training <- training[, features_select ]

cat("\nTraining set dimensions (after NAs removal): ", dim(training))

```

### 2.2. Remove low cardinality and non-informative features 

Some features has low cardinality (number of unique values): 

- The columns **new_window**, **user_name**, **cvtd_timestamp** were found not relevant for predicting the target. 
- Additional 7 non-informative features were removed from both training and test data sets. 

```{r remove_card, echo=TRUE}

# Identify low cardinality columns 
cardinality <- data.frame( as.table(sapply(training, function(x) length(unique(x)))) )
cardinality <- cardinality[ order(cardinality$Freq), ]

cat("Low cardinality features table:\n\n")
print( head(cardinality) )

# Remove other uninformative columns
training <- subset( training, select = -c( X, new_window, user_name, cvtd_timestamp, num_window, 
                                           raw_timestamp_part_1, raw_timestamp_part_2 ) )

# Filer out columns from testing data set 
problem_id = data.frame(problem_id = testing$problem_id )
testing <- subset( testing, select = names(subset(training, select=-c(classe))) )

cat("\n\nTraining set dimensions (after low cardinality removal): ", dim(training))
cat("\nTesting set dimensions (after low cardinality removal):  ", dim(testing))

```

### 2.3. Identify numeric and character features 

All the remaining  predictors were numerical features of either **float** or **integer** type.

```{r feature_types, echo=TRUE}

# Create lists of features 
features_num <- names(training)[ sapply(training, is.numeric) ]
features_int <- names(training)[ sapply(training, is.integer) ]
features_chr <- names(training)[ sapply(training, is.character) ]

n1 <- length(features_num)
n2 <- length(features_int)
n3 <- length(features_chr)

# Show the number of features 
cat("Number of numeric features:   ", n1 )
cat("\nNumber of integer features:   ", n2 )
cat("\nNumber of character features: ", n3 )

```

### 2.4. Correlation between numeric features 

Data exploration revealed that many features which are highly correlated with other variables: 

- A total of 22 features have correlation coefficient of 0.80 or higher. 

```{r feature_correlation, echo=TRUE}

# Correlation matrix 
corr_num <- cor( training[, features_num], use = "pairwise.complete.obs" )
for(i in 1:n1) { corr_num[i:n1, i] <- 0 }

# Convert to data frame 
corr_num <- data.frame( as.table(corr_num) ) 
corr_num <- subset( corr_num, Var1 != Var2 )
corr_num <- corr_num[ order(abs(corr_num$Freq), decreasing=TRUE), ]

cat("\nCorrelation matrix:\n\n")
print(head(corr_num, 10))

cat("\nNumber of feature pairs with correlation r > 0.8:\n", sum(corr_num$Freq > 0.8))

```

### 2.5. Transform Features using PCA 

To normalize and de-correlate the predictors, PCA transform has been used: 

- It was found that 27 PCA components explain the cumulative 95% of total variance. 
- The dimension of the training and testing data can be reduced after PCA transform. 
- The total number of features for machine learning might be reduced from 52 to 25. 

```{r test_PCA, echo=TRUE}

# Perform a PCA transform 
transform_pca <- prcomp( subset(training, select = -c(classe)), scale = TRUE )

pca_explained <- transform_pca$sdev^2 / sum(transform_pca$sdev^2)
pca_explained <- cumsum(pca_explained)

title <- paste("\nNumber of components that explain 95% variance:", which( pca_explained > 0.95 )[1])

plot(pca_explained, ylim=c(0,1), xlab = "Number of PCA Components", ylab = "Explained Variance",  main = title )
abline( h=0.95, col="red" )
abline( v=27, col="red" )
grid()

```

```{r transform_PCA, echo=TRUE}

# Create a PCA transformer
model_pca <- preProcess( training, method = "pca", thresh = 0.95, scale = TRUE )

# Transform training data 
train_pca <- predict( model_pca, subset(training, select=-c(classe)) )
train_pca <- cbind( train_pca, classe = training$classe )

# Transform testing data 
test_pca <- predict( model_pca, testing )

# Glimpse data
head(train_pca)

```

## 3. Build Classification Model 

Several different classification model were fit to the training data: 

- Learner 1 - K-Nearest Neighbors model; 
- Learner 2 - Random Forest model; 
- Learner 3 - Gradient Boost model; 

A 5-fold cross-validation approach was used to evaluate each model's performance. 

```{r enable_parallel, echo=FALSE}

# Enable parallel 
registerDoParallel(cores = 10)

```

### 3.1. Create and fit a KNN learner 

Main observations: 

- The number of neighbors $k$ hyperparameter was tuned using cross-valiation. 
- The classification accuracy increases sharply for $k$ values smaller than 7. 
- However, the minimum value for $k=5$ was taken to avoid overfitting on the training set. 
- Overall model accuracy on the training set is around 99%. 

```{r model_knn, echo=TRUE}

# Create and fit KNN model 
set.seed(123)
model_knn <- train(
  classe ~ .,
  data = training,
  preProcess = c("center","scale"),
  method = "knn",
  trControl = trainControl(method = "cv", number = 5, allowParallel = TRUE),
  tuneGrid = expand.grid(k = seq(5, 15, by = 2))
)

cat("Description of KNN model:\n\n")
print(model_knn)

cat("\n\nConfusion matrix for KNN model (training set):\n")
pred_train_knn <- predict( model_knn, training )
table( pred_train_knn, training$classe )

cat("\n\nAccuracy for KNN model (training set):\n")
round( mean(pred_train_knn == training$classe) , 4)

```

### 3.2.Create and fit a Random Forest learner

Main observations: 

- Hyperparameter **mtry** was optimized via cross-validation; 
- Overall model accuracy on the training set is around 98%. 

```{r model_rf, echo=TRUE}

# Create and fit RF model 
set.seed(123)
model_rf <- train(
  classe ~ .,
  data = training,
  method = "rf",
  trControl = trainControl(method = "cv", number = 5, allowParallel = TRUE),
  tuneGrid = expand.grid(mtry = 1:5),
  verbose = FALSE 
)

cat("Description of RF model:\n\n")
print(model_rf)

cat("\nConfusion matrix for RF model (training set):\n")
pred_train_rf <- predict( model_rf, training )
table( pred_train_rf, training$classe )

cat("\n\nAccuracy for RF model (training set):\n")
round( mean(pred_train_rf == training$classe) , 4)

```

### 3.3. Create and fit a Gradient Boost learner 

Main observations: 

- Hyperparameters **n.trees**, **interaction.depth**, and **shrinkage** were optimized via cross-validation; 
- Overall model accuracy on the training set is around 96%. 

```{r model_gbm, echo=TRUE}

# Hyperparameter grid 
grid_gbm <- expand.grid(
  n.trees = seq(100, 200, by = 50),
  interaction.depth = c(3, 5, 7),
  shrinkage = c(0.05, 0.1),
  n.minobsinnode = 10
)

# Create and fit GBM model 
set.seed(123)
model_gbm <- train(
  classe ~ ., 
  data = training, 
  method = "gbm", 
  trControl = trainControl(method = "cv", number = 5, allowParallel = TRUE),
  tuneGrid = grid_gbm,
  verbose = FALSE 
)

cat("Description of GBM model:\n\n")
print(model_gbm)

cat("\nConfusion matrix for GBM model (training set):\n")
pred_train_gbm <- predict( model_gbm, training )
table( pred_train_gbm, training$classe )

cat("\n\nAccuracy for GBM model (training set):\n")
round( mean(pred_train_gbm == training$classe) , 4)

```

### 3.4. Predict labels for the test set with a meta-model 

A meta-model was created using three individual learners: 

- The stacked model was trained using predicted values from three learners. 
- A C5.0 (rules based) model with default setting was used. 
- A small improvement in the accuracy on the training set was achieved. 

```{r train_meta, echo=TRUE}

# Combined individual learners' predictions and the target 
train_combined <- data.frame( p1 = pred_train_knn, p2 = pred_train_rf, p3 = pred_train_gbm, classe = training$classe )

# Create and fit a stacked meta-model 
model_meta <- train(
  classe ~ ., 
  method = "C5.0",
  trControl = trainControl(method = "cv", number = 5, allowParallel = TRUE),
  tuneGrid = expand.grid(model = "rules", trials = 10, winnow = TRUE),
  data = train_combined, 
  verbose = FALSE 
)

cat("Description of meta model:\n\n")
print(model_meta)

cat("Confusion matrix for meta model (training set):\n")
pred_train_meta <- predict( model_meta, train_combined )
table( pred_train_meta, training$classe )

cat("\n\nAccuracy for meta model (training set):\n")
round( mean(pred_train_meta == training$classe) , 4)

```

The meta-model was use to predict **classe** labels: 

- The stacked model predicted classes for the test set based on predictions from the three learners. 

```{r test_meta, echo=TRUE}

# Combined individual learners' predictions and the target 
test_combined <- data.frame(
  p1 = predict(model_knn, testing),
  p2 = predict(model_rf, testing),
  p3 = predict(model_gbm, testing)
)

cat("Predictions for the test set:\n\n")
test_predicted <- cbind( test_combined, problem_id = problem_id, predicted = predict( model_meta, test_combined ) )
print( test_predicted )

```

```{r disable_parallel, echo=FALSE}

# Close the parallel cluster 
stopImplicitCluster()

```

